---
title: "STAT571_Final Project"
author: "Shaolong Wu"
date: "4/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(data.table)
responses<-as.data.frame(fread("/Users/lorrywu/Desktop/571Datasets/Project_data/responses.csv",skip=1,header=FALSE))[,-1]

responses_test<-as.data.frame(fread("/Users/lorrywu/Desktop/571Datasets/Project_data/responses_test.csv",skip=1,header=FALSE))[,-1]

labels<-as.matrix(fread("/Users/lorrywu/Desktop/571Datasets/Project_data/labels.csv",skip=1,header=FALSE))[,-1]

labels_val<-as.matrix(fread("/Users/lorrywu/Desktop/571Datasets/Project_data/labels_val.csv",skip=1,header=FALSE))[,-1]

## ROI is categorical: the values are 1 to 8
roi <- read.csv("~/Desktop/571Datasets/Project_data/roi.csv")
roi_names <- as.matrix(fread("~/Desktop/571Datasets/Project_data/roi_names.csv",skip=1, header=FALSE))

stimuli_test<-as.matrix(fread("/Users/lorrywu/Desktop/571Datasets/Project_data/stimuli_test.csv",skip=1,header=FALSE))[,-1]

stimuli<-as.matrix(fread("/Users/lorrywu/Desktop/571Datasets/Project_data/stimuli.csv",skip=1,header=FALSE))[,-1]
```


Find the dimension of the data frame to be 1750 units by 8000 voxels of each unit, which is a stimuli picture.
```{r}
dim(responses)

nrow(responses[1])
v1<-responses[1]
# responses[1]$V2
# responses[2]$V3
v1$V2[1:5] ## the first five responding pixels to picture 1
```


now, we first want to specify the column names so that we can revoke them more easily:
```{r}
col.responses<-paste0("r_",as.character(c(1:ncol(responses))))
colnames(responses)<-col.responses

X <- responses
dim(X)

range(X$r_1)
```


```{r}
Y <- rep(NA,ncol(labels))
unique(labels[1,])
```

primitive EDA: 

1. find the range of all the variables
we realize that the response variable's value is not beyond [-1,1], but rather more expansive.

```{r}
## maximum and minimum of response values across the entire dataframe
max(X,na.rm=TRUE)
min(X,na.rm=TRUE)
```

plot the distribution of the response data: (this is a very wide data, so we may not)

symmetricity (Gausiann!!! 后面继续check)

we find that the data has indeed a mean of approximately 0, as the small discrepancy may be just some recorded error.

```{r}
mean(X$r_1)
mean(X$r_2)
mean(X$r_3)

### wildcard case, then have mean(X$r_1) to mean(X$r_1XXX)
### distribution of the response data in total 
hist(X$r_1, main="Histogram of 1st variable in response matrix",col="blue",xlab="response_1",ylab="number of occurence")
```


Some primitive correlation analysis:

Is there correlation between v1, v2?

```{r}
cor(X$r_1,X$r_2)
```

the whole correlation matrix of all responses:

```{r}
# cor(X, method = c("pearson", "kendall", "spearman"))
library("Hmisc")

X<-data.matrix(X) # coerce object into double
res2 <- rcorr(X)


## Extract the first five correlation coefficients
res2$r[1:5]
length(res2$r)
# Extract p-values

### see how many NAs are there
res2$P[1:5]
sum(is.na(res2$P)) ## 8428 NAs 数据本身没有NA, diagonal之类的process过程中出现NA

length(res2$P)
```
```{r}
hist(res2$r, col="blue", xlab="correlation coefficients of variables",ylab="number of variables",main="Histogram of Correlation Coefficients")
hist(res2$P, col="green",xlab="power of correlation of variables", ylab="P-values of all pairwise correlation",main="Histogram of Correlation Power")
```

Plot1: correlation coefficients distribution
Plot2: correlation power distribution
Plot3: dimension-reduction method (based on ROI names)
Plot4: boxplot of response levels vs ROI

```{r}
r1 <- responses[,which(roi==1)]
colnames(r1) <- paste0("V1_", as.character(c(1:ncol(r1))))
r2 <- responses[,which(roi==2)]
colnames(r2) <- paste0("V2_", as.character(c(1:ncol(r2))))
r3 <- responses[,which(roi==3)]
colnames(r3) <- paste0("V3_", as.character(c(1:ncol(r3))))
r4 <- responses[,which(roi==4)]
colnames(r4) <- paste0("V3A_", as.character(c(1:ncol(r4))))
r5 <- responses[,which(roi==5)]
colnames(r5) <- paste0("V3B_", as.character(c(1:ncol(r5))))
r6 <- responses[,which(roi==6)]
colnames(r6) <- paste0("V4_", as.character(c(1:ncol(r6))))
r7 <- responses[,which(roi==7)]
colnames(r7) <- paste0("LatOcc_", as.character(c(1:ncol(r7))))

X <- cbind(r1,r2,r3,r4,r5,r6,r7)
```

```{r}
pc <- prcomp(responses, scale=TRUE)
names(pc)
```

```{r}
### ROI represents the category in brain
boxplot_theme <- theme_bw() + theme(legend.position = "none",plot.margin = margin(t = 5, r = 50, b = 5, l = 0, unit = "pt"), axis.text.x = element_text(angle = -60, vjust = 0, hjust = 0))




ggplot(aes(x = forcats::fct_reorder(roi, -responses, .fun = median), y = responses, fill = roi)) + geom_boxplot() + xlab("ROI") +
ylab("Response level") + ggtitle("Response level by ROI") + boxplot_theme
```




